Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-01-22 draft.tex
[draft] change notation in detector-acceptance section
o explicitly state the assumptions
o replace eps by E in most places

%% Shortcuts %%
\newcommand{\aver}[1]{\langle #1 \rangle}
\newcommand{\dual}[1]{\tilde{#1}}
\newcommand{\est}[1]{\widehat{#1}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\nuvec}{\vec{\nu}}
\newcommand{\order}[1]{\mathcal{O}\left({#1}\right)}
\newcommand{\refapp}[1]{app.~(\ref{app:#1})}
\newcommand{\refeq}[1]{eq.~(\ref{eq:#1})}
\newcommand{\reffig}[1]{fig.~(\ref{fig:#1})}
\newcommand{\refsec}[1]{sec.~(\ref{sec:#1})}
\newcommand{\reftab}[1]{tab.~(\ref{tab:#1})}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\thvec}{\vec{\vartheta}}
\let\oldtheta\theta
\renewcommand{\theta}{\vartheta}
\let\eps\varepsilon
\newcommand{\vecest}[1]{\widehat{\vec{#1}}}
\newcommand{\wwhat}[1]{\widehat{\widehat{#1}}}

@@

\subsection{Recipe for the Determination of Detector Acceptance Effects}
\label{sec:systematics:acceptance}

The determination of a detector's performance to detect signal events
with accurate determination of the event's angles is generally a
difficult task. In the ideal case, one would have an explicit
probabilistic model of the detector acceptance effects and could thus
write down the full forward model from which the measured events
arise. In practice, this would require much more CPU time than
available, and one is therefore forced to simplify the model. The
standard approach is to generate the \emph{true} particle events
$\lbrace x_\text{t}^{(n)}\rbrace =
\lbrace(\vec{\nu}^{(n)}_\text{t},\vec\theta^{(n)}_\text{t})\rbrace$,
$n=1,\dots,N_\text{t}$, from a PDF assumed to describe the bare
physical process, and to propagate those particles through a detailed
simulation of the detector.  The observable traces that the particles
leave in the detector are fed into reconstruction algorithms resulting
in the \emph{detector} events $\lbrace x^{(n)}_\text{d}\rbrace =
\lbrace(\vec{\nu}^{(n)}_\text{d}, \vec\theta^{(n)}_\text{d})\rbrace$,
$n=1,\dots,N_\text{d}$.  In general, the distribution of the detected
events is
\begin{equation}
    P_\text{d}(x_\text{d}) = \int P_\text{t}(x_\text{t}) E(x_\text{d} | x_\text{t}) \rmdx{x_\text{t}}\,.
\end{equation}
Here $P_\text{t}$ is the probability distribution of the true events,
and the kernel $E(x_\text{d}| x_\text{t})$ parametrizes acceptance and
resolution effects. Note that $P_\text{d}$ is \emph{not} a PDF, since
it is not normalized,
\begin{equation}
    \int P_\text{d}(x) \rmdx{x} \neq 1\,.
\end{equation}
For the sake of clarity, we will restrict ourselves
to pure acceptance effects and suppress resolution effects in the remainder of this section; i.e.,
\begin{equation}
    E(x_\text{d}, x_\text{t}) \equiv \eps(x_\text{t}) \delta(x_\text{d}, x_\text{t})\,.
\end{equation}
We will refer to $\eps$ as the \emph{detector acceptance} function. In what follows,
we propose a systematic methods to unfold all effects of $\eps$ through MC simulations
and the method of moments.\\

We will now proceed with the explicit usage of a uniangular PDF with $x = \cos\theta$.\footnote{%
    The generalization of this section to multiangular PDFs is straighforward. It can be achieved
    by promoting $x$ to a vector, promoting the Legendre polynomials to products of independent polynomials
    or spherical harmonics, and promoting the indices $i,j,k,m$ to multi-indices.
}
Let us define the PDF in terms of the Legendre polymials (i.e., $f_k(x) \equiv P_k(x)$),
\begin{equation}
    P(x) \equiv \sum_k S_k f_k(x),
\end{equation}
where $k = 0, 1, \dots$ denotes the total angular momentum of the observables.
From the normalization of the PDF then follows $S_0 = 1/2$. From positivity of $P$ one
obtains $|S_k| \leq 1/2$ for angular momentum $k > 0$. More stringent relations between the
$S_k$ might hold, but are beyond the scope of this work. For latter use, we define
$\vec{S}^{(m)}_k \equiv 1/2 (\delta_{k,0} + \delta_{k,m})$, and we emphasize that
\begin{equation}
    P^{(m)}(x) \equiv P(x | \lbrace S^{(m)}_k \rbrace)
\end{equation}
is a PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
\begin{equation}
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x),\qquad \int_{-1}^{+1} \tilde{f}_k(x) f_l(x) \rmdx{x} = \delta_{k,l}\,.
\end{equation}

We now define the \emph{simulated raw} moments $\lbrace Q_{i}^{(m)}\rbrace$,
\begin{equation}
    Q_i^{(m)} \equiv \int_{-1}^{+1} \tilde{f}_i(x) P^{(m)}(x) \eps(x) \rmdx{x} = \frac{1}{2} \int_{-1}^{+1} \tilde{f}_i(x) \big[f_0(x) + f_m(x)\big] \eps(x) \rmdx{x}\,,
\end{equation}
which are instrumental to our recipe. MC estimators for these simulated moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$, where
\begin{equation}
    x_\text{d}^{(n,m)} \sim P_\eps^{(m)} \equiv \frac{P^{(m)} \eps}{R^{(m)}}\,.
\end{equation}
Here the normalization $R^{(m)}$ is chosen such that $\int P_\eps^{(m)}(x) \rmdx{x} = 1$.
We emphasize that $R^{(m)}$ can estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} / N_\text{t}$, where $N_\text{t}$ corresponds to the number of simulated true events.
The estimators then read
\begin{equation}
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}} \sum_n^{N_\text{d}} \tilde{f}_i(x_\text{d}^{(n,m)})
\end{equation}

Linearity of the integral over $x$ and convergence of the expansion of $P_\eps^{(m)}$ in terms Legendre polynomials ensures that
\begin{equation}
    Q_i^{(m)} = \sum_j M_{ij} S_j^{(m)}\,.
\end{equation}
We call the matrix $M_{ij}$ the \emph{unfolding matrix}, which is specific to our decay at hand. Given our definition of
$S_k^{(m)}$ it can easily be calculated as
\begin{equation}
    M_{ij} = \begin{cases}
        2 Q_i^{(0)}                          & j = 0\,,\\
        2\left(Q_i^{(j)} - Q_i^{(0)}\right)  & j \neq 0\,,\\
    \end{cases}
\end{equation}
and its MC estimator $\est{M}_{ij}$  can be obtained through the replacements $Q_i^{(m)} \to \est{Q}_i^{(m)}$.\\

In order to finally extract the angular observables from data, we use the \emph{measured raw moments}. Their MC estimator
$\lbrace \est{Q}_i\rbrace$ -- based on the \emph{detected events} $x^{(n)}$, $n=1,\dots,N$ -- reads
\begin{equation}
    \est{Q}_i \equiv R \frac{1}{N} \sum_n^N \tilde{f}_i(x^{(n)})\,.
\end{equation}
Here $R \equiv \int P(x) \eps(x)$ is an a priori unknown normalization that also occurs in the determination of the branching
ratio of the specific decay. We then obtain MC estimators of the angular observables via
\begin{equation}
    \label{eq:estS}
    \est{S}_i \equiv \sum_{i} \left(\hat{M}^{-1}\right)_{ij} \est{Q}_j\,.
\end{equation}
The precise value of $R$ depends on the a priori unknown angular observables $\lbrace S_i\rbrace$.
However, ratios and other suitable combinations of the estimators $\lbrace \est{S}_i\rbrace$ are independent
of $R$. Conversely, $R$ can be extracted from the self-consistency condition\footnote{This condition
is process dependent; see \refeq{recipe:lambdabtolambdall} and \refeq{recipe:btokstarll}.}
$\hat{S}_0 = \frac{1}{2}$. Since $R$ is relevant in the determination of the branching ratio, it is therefore useful
to carry out that determination and the extraction of the angular observables in a combined analysis.

\danny{
We note that the set $\lbrace \est{S}_k\rbrace$ as determined from \refeq{estS} does not fully correspond to the
set $\lbrace S_k\rbrace$ for an arbitrary $\eps$. Taking again the decay $B\to K\ell^+\ell^-$
as an example, we have $\dim S = 3$. Assuming an expansion of the detector acceptance $\eps$
in terms of Legendre polynomial up to order $L$, we obtain $\dim \est{Q} = \dim \est{S} = \dim S + L$.
The MC estimators then have the following structure:
\begin{equation}
    \lbrace \est{S}_k \rbrace = \lbrace \underbrace{\est{S}_0, \dots, \est{S}_{\dim S}}_\text{physical}, \underbrace{\est{S}_{\dim S + 1}, \dots, \est{S}_{\dim Q}}_\text{``superfluous''}\rbrace\,.
\end{equation}
The method is consistent as long as the MC estimators for the ``superfluous'' observables with indices $\dim S + 1$ to $\dim Q$
are compatible with zero. \footnote{This holds only for the physical model of uniangular decay distributions. For some decays, these superfluous observables
correspond to higher partial waves that have been suppressed in the physical model; see \refapp{btokstarll} for such a case.}
Since $L$ depends on the construction of the particle detector at hand, it remains for
the experimental collaborations to determine the maximum order of such an expansion.
}

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates. While
considerations of any detector simulation is beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuarcy. We find for a uniangular distribution and a total of $6\cdot 10^6$ MC
samples (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$) that the error on the mean is
smaller than $6\cdot 10^{-4}$. Let us conclude this section by commenting that the unfolding
matrix is universal in a sense. Specifically, it can be reused in analyses with a similar underlying decay.
Therefore, computing resources spent on improving its accuracy are therefore not wasted.

>>>>

%% Shortcuts %%
\newcommand{\aver}[1]{\langle #1 \rangle}
\newcommand{\dual}[1]{\tilde{#1}}
\newcommand{\est}[1]{\widehat{#1}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\nuvec}{\vec{\nu}}
\newcommand{\order}[1]{\mathcal{O}\left({#1}\right)}
\newcommand{\refapp}[1]{app.~(\ref{app:#1})}
\newcommand{\refeq}[1]{eq.~(\ref{eq:#1})}
\newcommand{\reffig}[1]{fig.~(\ref{fig:#1})}
\newcommand{\refsec}[1]{sec.~(\ref{sec:#1})}
\newcommand{\reftab}[1]{tab.~(\ref{tab:#1})}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\subd}{_{\text{d}}}
\newcommand{\subt}{_{\text{t}}}
\newcommand{\thvec}{\vec{\vartheta}}
\let\oldtheta\theta
\renewcommand{\theta}{\vartheta}
\let\eps\varepsilon
\newcommand{\vecest}[1]{\widehat{\vec{#1}}}
\newcommand{\wwhat}[1]{\widehat{\widehat{#1}}}
\newcommand{\xd}{x\subd}
\newcommand{\xt}{x\subt}
\newcommand{\Ekernel}{E(\xd|\xt)}

@@

\subsection{Recipe for the Determination of Detector Effects}
\label{sec:systematics:acceptance}

The determination of a detector's performance to detect signal events
with accurate determination of the event's angles is generally a
difficult task. In the ideal case, one would have an explicit
probabilistic model of the detector acceptance effects and could thus
write down the full forward model from which the measured events
arise. In practice, this would require much more CPU time than
available, and one is therefore forced to simplify the model. The
standard approach is to generate the \emph{true} particle events
$\lbrace x_\text{t}^{(n)}\rbrace =
\lbrace(\vec{\nu}^{(n)}_\text{t},\vec\theta^{(n)}_\text{t})\rbrace$,
$n=1,\dots,N_\text{t}$, from a PDF assumed to describe the bare
physical process, and to propagate those particles through a detailed
simulation of the detector.  The observable traces that the particles
leave in the detector are fed into reconstruction algorithms resulting
in the \emph{detector} events $\lbrace x^{(n)}_\text{d}\rbrace =
\lbrace(\vec{\nu}^{(n)}_\text{d}, \vec\theta^{(n)}_\text{d})\rbrace$,
$n=1,\dots,N_\text{d}$.  In general, the distribution of the detected
events is
    P_\text{d}(x_\text{d}) = \frac{1}{R} \int P_\text{t}(x_\text{t}) E(x_\text{d} | x_\text{t}) \rmdx{x_\text{t}}\,.
Here $P_\text{t}$ is the probability distribution of the true events,
\fred{the normalization constant $R$ is given by
    R \equiv \int \int P_\text{t}(x_\text{t}) E(x_\text{d} | x_\text{t}) \rmdx{x_\text{t}} \rmdx{x_\text{d}}\,,
and the kernel $E(x_\text{d}| x_\text{t})$ parametrizes acceptance and
resolution effects. The kernel is usually decomposed as
  \Ekernel = \eps(\xt) P(\xd|\xt),
where the PDF $P(\xd|\xt)$ includes the resolution effects and the
unnormalized density $\eps(\xt)$ is the \emph{detector acceptance}
function. Perfect resolution corresponds to
  \label{eq:def-R}
  P(\xd|\xt) = \delta(\xd - \xt) \,.
}
% For the sake
% of clarity, we will restrict ourselves to pure acceptance effects and
% suppress resolution effects in the remainder of this section; i.e.,
% \begin{equation}
%     E(x_\text{d}, x_\text{t}) \equiv \eps(x_\text{t}) \delta(x_\text{d}, x_\text{t})\,.
% \end{equation}
% We will refer to $\eps$ as the \emph{detector acceptance} function.
In what follows, we propose a systematic method to unfold all effects
of \fred{$\Ekernel$} through MC simulations and the method of moments
\fred{assuming that $\Ekernel$ is independent of the angular
  observables, and that it is possible to expand $\Ekernel$ --- at
  least formally --- in Legendre polymials}.  For illustration, we
proceed with the explicit example of a uniangular\footnote{%
  The generalization of this section to multiangular PDFs is
  straighforward. It can be achieved by promoting $x$ to a vector,
  promoting the Legendre polynomials to products of independent
  polynomials or spherical harmonics, and promoting the indices
  $i,j,k,m$ to multi-indices.  } PDF with $x = \cos\theta$. Let us
define the PDF in terms of the Legendre polymials (i.e., $f_k(x)
\equiv P_k(x)$) \fred{and angular observables $\lbrace S_k \rbrace$ as
    P\subt(\xt) \equiv P\subt(\xt | \lbrace S_k \rbrace) =   \sum_k S_k f_k(\xt),
  \end{equation}}
where $k = 0, 1, \dots$ denotes the total angular momentum of the observables \fred{adjust ``angular momentum'' to language at the end of the section}.
\fred{Normalization of $P\subt$ is equivalent to choosing} $S_0 = 1/2$. \fred{Requiring $P\subt(\xt) \ge 0 \,\forall \xt$ implies} $|S_k| \leq 1/2$ for angular momentum \fred{$k$ just an index, can we remove ``angular momentum''?} $k > 0$. More stringent relations between the
$S_k$ might hold, but are beyond the scope of this work. For later use, we define
\fred{
    \label{eq:def-Sk}
  \vec{S}^{(m)}_k \equiv
  \begin{cases}
    1/2 \delta_{k,0}, & m=0\\
    1/2 (\delta_{k,0} + \delta_{k,m}), & m>0 \,,
  \end{cases}
\end{equation}}
and \fred{note} that
    P^{(m)}(\fred{\xt}) \equiv P(\fred{\xt} | \lbrace S^{(m)}_k \rbrace)
is a \fred{valid} PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x),\qquad \int_{-1}^{+1} \tilde{f}_k(x) f_l(x) \rmdx{x} = \delta_{k,l}\,.
We now define the \emph{simulated raw} moments $\lbrace Q_{i}^{(m)}\rbrace$ as
    Q_i^{(m)} \equiv \fred{\int \int \tilde{f}_i(\xd) P^{(m)}(\xt) \Ekernel \rmdx{\xt} \rmdx{\xd},}
which are instrumental to our recipe. \fred{Monte Carlo} estimators of these moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$, where
    x_\text{d}^{(n,m)} \sim \fred{P_E^{(m)}(\xd) \equiv \frac{1}{R^{(m)}} \int P^{(m)}(\xt) \Ekernel \rmdx{\xt}}\,.
  \fred{In words, for each $m$ it is required to generate events from
    a toy physical distribution in which $S_0 = S_m=1/2$ and all other
    observables are set to zero, and to propagate those events through
    a detector simulation.} The normalization $R^{(m)}$ is chosen such
  that \fred{$\int P_E^{(m)}(\xd) \rmdx{\xd} = 1$}.  We emphasize that
  $R^{(m)}$ can be estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} /
  N_\text{t}$, where $N_\text{t}$ corresponds to the number of
  simulated true events.  The estimators then read
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}} \sum_n^{N_\text{d}} \tilde{f}_i(x_\text{d}^{(n,m)})

Linearity of the integral over $x$ and convergence of the expansion of
$P_E^{(m)}$ in terms of Legendre polynomials ensures that
\begin{equation}
    Q_i^{(m)} = \sum_j M_{ij} S_j^{(m)}\,.
\end{equation}
We call the matrix $M_{ij}$ the \emph{unfolding matrix}, which is
specific to the decay at hand. Given our definition of $S_k^{(m)}$ in
\fred{\refeq{def-Sk} it is easy to see that}
\begin{equation}
    M_{ij} = \begin{cases}
        2 Q_i^{(0)}                          & j = 0\,,\\
        2\left(Q_i^{(j)} - Q_i^{(0)}\right)  & j \neq 0\,,\\
    \end{cases}
\end{equation}
and its MC estimator $\est{M}_{ij}$  can be obtained through the replacements $Q_i^{(m)} \to \est{Q}_i^{(m)}$.\\

In order to finally extract the angular observables from data, we use the \emph{measured raw moments}. Their MC estimator
$\lbrace \est{Q}_i\rbrace$ -- based on the \emph{detected events} $x^{(n)}$, $n=1,\dots,N$ -- reads
\begin{equation}
    \est{Q}_i \equiv R \frac{1}{N} \sum_n^N \tilde{f}_i(x^{(n)})\,.
\end{equation}
We then obtain MC estimators of the angular observables via
\begin{equation}
    \label{eq:estS}
    \est{S}_i \equiv \sum_{i} \left(\hat{M}^{-1}\right)_{ij} \est{Q}_j\,.
\end{equation}
The precise value of $R$ (see \refeq{def-R}) depends on the a priori
unknown angular observables $\lbrace S_i\rbrace$.  However, ratios and
other suitable combinations of the estimators $\lbrace
\est{S}_i\rbrace$ are independent of $R$. Conversely, $R$ can be
extracted from the self-consistency condition\footnote{This condition
  is process dependent; see \refeq{recipe:lambdabtolambdall} and
  \refeq{recipe:btokstarll}.}  $\hat{S}_0 = \frac{1}{2}$. Since $R$ is
relevant in the determination of the branching ratio, it is therefore
useful to carry out that determination and the extraction of the
angular observables in a combined analysis.

\danny{
We note that the set $\lbrace \est{S}_k\rbrace$ as determined from \refeq{estS} does not fully correspond to the
set $\lbrace S_k\rbrace$ for an arbitrary $\eps$. Taking again the decay $B\to K\ell^+\ell^-$
as an example, we have $\dim S = 3$. Assuming an expansion of the detector acceptance $\eps$
in terms of Legendre polynomial up to order $L$, we obtain $\dim \est{Q} = \dim \est{S} = \dim S + L$.
The MC estimators then have the following structure:
\begin{equation}
    \lbrace \est{S}_k \rbrace = \lbrace \underbrace{\est{S}_0, \dots, \est{S}_{\dim S}}_\text{physical}, \underbrace{\est{S}_{\dim S + 1}, \dots, \est{S}_{\dim Q}}_\text{``superfluous''}\rbrace\,.
\end{equation}
The method is consistent as long as the MC estimators for the ``superfluous'' observables with indices $\dim S + 1$ to $\dim Q$
are compatible with zero. \footnote{This holds only for the physical model of uniangular decay distributions. For some decays, these superfluous observables
correspond to higher partial waves that have been suppressed in the physical model; see \refapp{btokstarll} for such a case.}
Since $L$ depends on the \fred{setup} of the particle detector \fred{under consideration}, it remains for
the experimental collaborations to determine the maximum order of such an expansion.
}

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates. While
considerations of any detector simulation is beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. \fred{For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find} that the error on the mean is
smaller than $6\cdot 10^{-4}$. Let us conclude this section by commenting that the unfolding
matrix is universal in a sense. Specifically, it can be reused in analyses with a similar underlying decay.
Therefore, computing resources spent on improving its accuracy are therefore not wasted.
